---
title: "Assignment 3: Data Exploration"
author: Janice Ye
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---


## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.


## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}

#checking working directory
getwd()

#installing packages
install.packages("tidyverse")
install.packages("lubridate")

#loading packages
library(tidyverse)
library(lubridate)

#upload datasets
neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv",stringsAsFactors = TRUE)
litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",stringsAsFactors = TRUE)
```


## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: People might be interested in the ecotoxicology of neonicotinoids on insects for research purposes. For example, people might want to know the impact and harm of each insecticides on crops adn human health.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: People can use these data to research on forest carbon accounting because woody debris is crucial to ecosystem due to its important role in carbon budget and nutrient cycling.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. litter trap pair (one elevated trap and one ground trap) is deployed for every 400 m2 plot area, resul􏰁ng in 1-4 trap pairs per plot. In some cases, available space, plot spacing requirements, and/or the tower airshed size re- stricts the number of plots that can be sampled for li􏰀er below 20 (forested) or 30 (low-stature).
 2. In sites with forested tower airsheds, the li􏰀er sampling is targeted to take place in 20 40m x 40m plots.
 3. In sites with low-statured vegeta􏰁on over the tower airsheds, li􏰀er sampling is targeted to take place in 4 40m x 40m tower plots (to accommodate co-located soil sampling) plus 26 20m x 20m plots.


## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
dim(neonics)
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}
summary(neonics$Effect)
```

> Answer:the most common effects that are studied are population and mortality. These effects can be used to identify the relationship between mortality rates and pesticides.

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r}
summary(neonics$Species.Common.Name)
```

> Answer:the six most commonly studied specides in the dataset are honey bee, parasitic wasp, buff tailed bumblebee, carniolan honey bee, bumble bee, and Italian honeybee. They all look alike and they are all insects. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r}
class(neonics$Conc.1..Author.)
```

> Answer:It is a factor because it is a categorical value instead of numeric value.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
library(ggplot2)

ggplot(neonics, aes(x = Publication.Year)) + 
  geom_freqpoly()
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(neonics) +
  geom_freqpoly(aes(x = Publication.Year)) +
  geom_freqpoly(aes(x = Publication.Year), subset(neonics, Test.Location == "Field artificial"), color = "red") +
  geom_freqpoly(aes(x = Publication.Year), subset(neonics, Test.Location == "Field natural"), color = "green") +
  geom_freqpoly(aes(x = Publication.Year), subset(neonics, Test.Location == "Field undeterminable"), color = "blue") +
  geom_freqpoly(aes(x = Publication.Year), subset(neonics, Test.Location == "Lab"), color = "orange")
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:The most common test locations are in lab and in natural field. They do differ over time. Before 2000, natural fields were more commonly used than lab. After 2000, lab became generally more common. One exception is between 2000 and 2010 where lab and natural fields were both in favor.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
ggplot(neonics) +
  geom_bar(aes(x = Endpoint)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

> Answer:The two most common endpoints are NOEL and LOEL. NOEL is no-observable-effect-level. It is the highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test. LOEL is the lowest-observable-effect-level. It is the lowest dose (concentration) producing effects that were significantly different from responses of controls.


## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
#determine the class of collectDate
class(litter$collectDate)

#change it to a date
litter$collectDate <- as.Date(litter$collectDate, format = "%Y-%m-%d") 
class(litter$collectDate)

#litter sampled in Aug 2018
unique(litter$collectDate, 2018-08 == FALSE)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
summary(litter)
unique(litter)
```

> Answer: There are 52 plots if we use the 'unique' function. The 'summary' fuction will result in much more plots compared to using the 'unique' function because 'summary' function did not eliminated duplicates.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
ggplot(litter) +
  geom_bar(aes(x = functionalGroup))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
#boxplot
ggplot(litter) + 
  geom_boxplot(aes(x = functionalGroup, y = dryMass))

#violin plot
ggplot(litter, aes(x = functionalGroup, y = dryMass)) +
  geom_violin()
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: Boxplot is more effective than the violin plot becasue the data is nearly identical in this case which will make the violin plot look aligned rather than having a density curve.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles tend to have the highest biomass at these sites.
