---
title: "Assignment 2: Coding Basics"
author: "Janice Ye"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
  seq(1,30,3)   #generating a sequence from 1 to 30 by 3
  seq_by_3 <- seq(1,30,3)   #naming the sequence "seq_by_3"

#2. 
  mean(seq_by_3)  #finding the mean of the sequence
  median(seq_by_3)  #finding the median of the sequence

#3. 
  mean(seq_by_3) > median(seq_by_3)   #asking a conditional statement if the mean is greater than the median
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#name is a character vector
name <- c("Amy", "Bob", "Carl", "David")
#score is a numeric vector
score <- c(60,70,40,80)
#pass is a logical vector
pass <- score >= 50

test_score <- data.frame(name, score, pass)

test_score
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Matrix is homogeneous while data frame is heterogeneous. This means that matrix can only store one type of data whereas data frame can store different types of data. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#ifelse
pass_fail <- ifelse(score >= 50, "PASS","FAIL")
print(pass_fail)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: ‘ifelse’ would work instead of "if" and "else" because score is a vector. "Ifelse" is for vector whereas "if" and "else" is for scalar.
