---
title: "Assignment 10: Data Scraping"
author: "Janice Ye"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)
library(here) 

here()

# Set theme
mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black",
                                 hjust = 0.5), 
        legend.position = "top")
theme_set(mytheme)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
website <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
website
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
water.system.name <- website %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water.system.name

PWSID <- website %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
PWSID

ownership <- website %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

max.day.use <- website %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
max.day.use
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 
Month <- c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec")

withdrawal.df <- data.frame("Month" = Month,
                            "Year" = rep(2022, 12),
                            "Max.Day.Use" = as.numeric(max.day.use))

withdrawal.df <- withdrawal.df %>%
  mutate(System.Name =!!water.system.name,
         PWSID =!!PWSID,
         Ownership =!!ownership,
         Date = my(paste(Month,"-",Year))) %>%
  select(Date, System.Name, PWSID, Ownership, Max.Day.Use) %>%
  arrange(Date)

#5 
withdrawal.plot <- ggplot(withdrawal.df, aes(x = Date, y = Max.Day.Use)) + 
  geom_line() + 
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = paste("2022 Maximum Day Use Withdrawal"),
       subtitle = water.system.name,
       y = "Withdrawal (mgd)",
       x = "Date") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
withdrawal.plot
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
scrape <- function(pwsid, year) {
  scrape.website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', 
                                     pwsid, '&', 'year=', year))
  
  water.system.name.id <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  PWSID.id <- 'td tr:nth-child(1) td:nth-child(5)'
  ownership.id <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  max.day.use.id <- 'th~ td+ td'
  
  water.system.name <- scrape.website %>% html_nodes(water.system.name.id) %>% html_text()
  PWSID <- scrape.website %>% html_nodes(PWSID.id) %>% html_text()
  ownership <- scrape.website %>% html_nodes(ownership.id) %>% html_text()
  max.day.use <- scrape.website %>% html_nodes(max.day.use.id) %>% html_text()

  Month <- c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec")

  withdrawal.df <- data.frame("Month" = Month,
                              "Year" = rep(year, 12),
                              "Max.Day.Use" = as.numeric(max.day.use))

  withdrawal.df <- withdrawal.df %>%
    mutate(System.Name =!!water.system.name,
          PWSID =!!PWSID,
          Ownership =!!ownership,
          Date = my(paste(Month,"-",Year))) %>%
    select(Date, System.Name, PWSID, Ownership, Max.Day.Use) %>%
    arrange(Date)
  
    Sys.sleep(1) 
    
    return(withdrawal.df)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
withdrawal.2015 <- scrape('03-32-010', 2015)
view(withdrawal.2015)

withdrawal.2015.plot <- ggplot(withdrawal.2015, aes(x = Date, y = Max.Day.Use)) + 
  geom_line() + 
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = paste("2015 Maximum Day Use Withdrawal"),
       subtitle = water.system.name,
       y = "Withdrawal (mgd)",
       x = "Date") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
withdrawal.2015.plot
```


8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
asheville.2015 <- scrape('01-11-010', 2015)
view(withdrawal.2015)

combined.df <- rbind(asheville.2015, withdrawal.2015)

combined.plot <- ggplot(combined.df, aes(x = Date, y = Max.Day.Use)) + 
  geom_line(aes(color = System.Name)) + 
  labs(title = paste("2015 Maximum Day Use Withdrawal"),
       y = "Withdrawal (mgd)",
       x = "Date") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
combined.plot
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
years = rep(2010:2021)
A.pwsid = '01-11-010'

asheville.years <- map(years, scrape, pwsid = A.pwsid)
asheville.year <- bind_rows(asheville.years)

asheville.plot <- ggplot(asheville.year, aes(x = Date, y = Max.Day.Use)) + 
  geom_line() + 
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = paste("Asheville Maximum Day Use Withdrawal from 2010 to 2021"),
       y = "Withdrawal (mgd)",
       x = "Date") +
  theme(plot.title = element_text(hjust = 0.5))
asheville.plot
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer: Yes, Asheville does have a trend in water usage over time. We can see that the usage was gradually decreasing from 2010 to 2015, then start to increase afterwards.
